---
title: "DATA 606 Data Project Proposal"
author: "Vic Chan"
output: pdf_document
---

### Data Preparation

```{r setup, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
# load data
dataset = read.csv('https://www.fueleconomy.gov/feg/epadata/vehicles.csv')

# Filtering out all options that are unnecessary

library(tidyr)
library(dplyr)

dataset = dataset %>%
  filter(fuelType == 'Regular' |  fuelType == 'Premium' | fuelType == 'Midgrade') %>%
  select(city08, highway08, cylinders, displ, drive, fuelType, make, model, year, trany)
```


### Research question 

**You should phrase your research question in a way that matches up with the scope of inference your dataset allows for.**

Have fuel economy increased throughout the years for gas engine cars or have they peaked? This means that we will be needing to filter out diesel, electric, and hybrid vehicles as they will impact the study.

### Cases 

**What are the cases, and how many are there?**

Each row represents a vehicle make and model. In the dataset there are 43,418 different vehicle make and models spanning from 1984 to 2021.

### Data collection 

**Describe the method of data collection.**

The data is collected from [The US Department of Energy](https://www.fueleconomy.gov/feg/download.shtml) and includes all make and models of vehicles from 1984 to 2021. 

### Type of study 

**What type of study is this (observational/experiment)?**

This is a observational study

### Data Source 

**If you collected the data, state self-collected. If not, provide a citation/link.**

U.S. Department of Energy, Energy Information Administration, Independent Statistics & Analysis. (2021, March 25).  Fuel Economy Data.  Retrieved from https://www.fueleconomy.gov/feg/download.shtml


### Dependent Variable

**What is the response variable? Is it quantitative or qualitative?**

The response variable is miles per gallon city/highway and it is quantitative.

### Independent Variable

**You should have two independent variables, one quantitative and one qualitative.**

Quantitative - Cylinders
Qualitative - Drive Train
Qualitative - Transmission Type
Quantitative - Displacement
Quantitative - Year

### Relevant summary statistics 

**Provide summary statistics for each the variables. Also include appropriate visualizations related to your research question (e.g. scatter plot, boxplots, etc). This step requires the use of R, hence a code chunk is provided below. Insert more code chunks as needed.**

```{r}
library(ggplot2)

dataset %>%
  ggplot(aes(x=city08)) + geom_histogram(bins = 50)
```
```{r}
dataset %>%
  ggplot(aes(x=highway08)) + geom_histogram(bins = 27)
```
```{r}
library(GGally)

ggpairs(dataset[,c("city08", "highway08", "displ", "cylinders")])
```

